---
title: "C14MC"
author: "Melanie Smith"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Import Counts and Sample Data

### Import Count Data 

```{r message=FALSE, warning=FALSE, error=FALSE}

# load required libraries
library(plyr)
library(reshape2)
library(tidyverse)
library(magrittr)
library(pheatmap)
library(edgeR)


# import placenta miRNA counts table from VM
countsPlacenta <- read.table(("~/placentaBcbio/placenta/final/2017-05-23_placenta/counts_mirna.tsv"), 
                             sep = "\t", 
                             header = T, 
                             row.names = "mirna")

# remove the "_R1" and replace "Q" with "PAC" from the column names (eg Q006_S1_R1 becomes PAC006 to match the 'samples' data.frame) in the 'counts' data.frame
names(countsPlacenta) = sub("_R1","",names(countsPlacenta)) 
names(countsPlacenta) = sub("Q","PAC",names(countsPlacenta))
names(countsPlacenta) = sub("_S.*","",names(countsPlacenta))
# correct sample name mislabeled by QIAGEN
names(countsPlacenta) = sub("PAC033","PAC038",names(countsPlacenta))
# Reorder data.frame by columns
countsPlacenta <- countsPlacenta[,order(colnames(countsPlacenta),decreasing = FALSE)]

# Import plasma miRNA counts table from VM
countsPlasma <- read.table(("~/plasmaBcbio/plasma/final/2017-05-22_plasma/counts_mirna.tsv"), 
                sep = "\t",
                header = T, 
                row.names = "mirna")

# remove the "S#_R1_001" from the column names (eg PAC001_S1_R1_001 becomes PAC001 to match the 'samples' data.frame) in the 'counts' data.frame
names(countsPlasma) = sub("_R1_001","",names(countsPlasma))
names(countsPlasma) = sub("_S.*","",names(countsPlasma))

# meta data for miRNA IDs - includes chromosome name and location
mirbase_db <- readRDS("~/Bcbioplasma/mirbase_db.rds")

```

### Import Sample Meta Data 

```{r message=FALSE, warning=FALSE, error=FALSE}

# import the samples data from VM 
samples <- readr::read_delim("~/placenta_miRNA.tsv", delim = "\t") %>% 
  dplyr::select(-starts_with("X"))

samples <- samples[,c(1, 12:16)] %>%
  as.data.frame() %>%
  set_colnames(c("samplename", "maternalAge", "gestation", "BMI", "smoker", "ethnicity")) %>% 
  mutate(samplename = paste0("PAC",stringr::str_pad(samplename, width = 3, pad = "0")))

# Remove the "+" from column 4 "gestation" nb: escape special character using \\
samples$gestation <- gsub("\\+", "", samples$gestation)

# reassign object classes
samples$gestation <- as.integer(samples$gestation)

# add new column named "group" into samples data.frame 
samples$group <- rep(NA, 48)

# add new group 'Trimester 1 or 2' factor to each row based on gestation
# Trimester 1 = 6-13 wks, Trimester 2 = 14-26 wks
samples$group <- ifelse(samples$gestation <= 13, "Trimester_1", 
                        ifelse(samples$gestation >= 14, "Trimester_2", NA))

# add new oxygenation factor to each row based on gestation
# hypoxic = 6-9 wks, oxic = 10-26 wks
samples$oxygenation <- ifelse(samples$gestation <= 9, "hypoxic", 
                        ifelse(samples$gestation >= 10, "oxygenated", NA))


samples <- as.data.frame(samples)

```

### Reasign object classes

```{r message=FALSE, warning=FALSE, error=FALSE}

# group as factor
samples$group <- as.factor(samples$group)

# group as factor
samples$oxygenation <- as.factor(samples$oxygenation) 


```

### Import and add fetal sex data to Sample Meta Data 

```{r message=FALSE, warning=FALSE, error=FALSE}

# Add fetal sex to samples df
fetalSex <- read.csv("~/Bcbioplasma/fetalSex.csv", header = TRUE) %>%
  dplyr::select(-matches("gestation")) %>%
  mutate(samplename = paste0("PAC", stringr::str_pad(samplename, width = 3, pad = "0")))

fetalSex$Sex <- ifelse(fetalSex$fetalSex == "XX" , "Female", 
                       ifelse(fetalSex$fetalSex == "XY", "Male", NA))

samples <- left_join(samples, fetalSex, by = "samplename")

```

# Filter and Normalise placenta counts

## Filter sequencing noise - Placenta

Here any individual observation with less than five counts is considered to be sequencing noise and is reduced to zero and removed by filtering.  
  * Counts less than 5 reduced to 0
  * Logic check and record how many miRs will be filtered
  * Filter to remove miRs with zero counts in all samples

```{r warning=FALSE, message=FALSE}

# exclude any individual count less than 5
countsPlacenta[countsPlacenta < 5] <- 0

# check to establish if any miRNA have no expression across all samples
table(rowSums(countsPlacenta == 0) == 48)

# remove miRNAs with zero counts in all samples
countsPlacenta <- countsPlacenta[ rowSums(countsPlacenta) != 0, ]

```

## Establish DGEList object - Placenta

The counts and samples data can now be combined into a single list object for use in differential expression analysis downstream.

```{r message=FALSE, warning=FALSE}

# new df of miRNA names for DGEList object
genes <- as.character(rownames(countsPlacenta)) %>%
  as.data.frame() %>%
  set_colnames("SYMBOL") 

# new DGEList object for differential expression analysis
DGEList_placenta <- DGEList(counts = countsPlacenta, samples = samples, genes = genes)


```

## TMM normalisation of libraries - Placenta

Before differential expression analysis the sample libraries need to be normalised to account for differences in initial library size.     
Normalising the libraries allows for the direct comparison between samples.   
Here the Trimmed Mean of M Values method is used.   
Prior to normalisation the placenta libraries are within a similar range, however the mean of counts is small and centered around 0-1 CPM (log2)   

```{r warning=FALSE}

# calculate normalisation factors and apply to the DGEList object
DGEList_placenta <- calcNormFactors(DGEList_placenta, method = "TMM")

```

## Filter biological noise - Placenta 

Once the libraries are normalised we can filter for biological noise - i.e. miRNAs that are present in such small numbers that they are not biologically relevant. Here a count of less than 1 CPM in 7 or more samples is considered biologically relevant for this analysis.  

```{r warning=FALSE, error=FALSE}

# new df of unfiltered cpm for the reduced DGEList
rawCpm_placenta <- cpm(DGEList_placenta)

# new df of unfiltered log 2 cpm for the reduced DGEList
rawlcpm_placenta <- cpm(DGEList_placenta, log = TRUE)

# remove low expressed genes logical eg cpm >1 in >=20 samples (7 samples in the smallest group)
a <- 1 # CPM for filtering
b <- 7 # number of samples for filtering

# remove low expressed miRNA 
keep.exprs <- rowSums(rawCpm_placenta > a) >= b
DGEList_placenta <- DGEList_placenta[keep.exprs,, keep.lib.sizes = FALSE]

```

# Filter and Normalise plasma counts

## Filter sequencing noise - Plasma   

Here any individual observation with less than 5 counts is considered to be sequencing noise and is reduced to zero and removed by filtering.  
  * Counts less than 5 reduced to 0
  * Logic check and record how many miRs will be filtered
  * Filter to remove miRs with zero counts in all samples

```{r warning=FALSE, message=FALSE, error=FALSE}

# exclude any individual count less or equal to than 5
countsPlasma[countsPlasma <= 5] <- 0

# check to establish if any miRNA have no expression across all samples
table(rowSums(countsPlasma == 0) == 48)

# remove miRNAs with zero counts in all samples
countsPlasma <- countsPlasma[ rowSums(countsPlasma) != 0, ]

```

## Establish DGEList object - Plasma

The counts and samples data can now be combined into a single list object for use in differential expression analysis downstream.

```{r warning=FALSE, message=FALSE, error=FALSE}

# assign object to DGEList
DGEList_plasma <- DGEList(counts = countsPlasma, samples = samples)

```

## TMM normalisation of libraries

Before differential expression analysis the sample libraries need to be normalised to account for differences in initial library size.     
Normalising the libraries allows for the direct comparison between samples.   
Here the Trimmed Mean of M Values method is used.   

```{r warning=FALSE, message=FALSE, error=FALSE}

# calculate normalisation factors and apply to the DGEList object
DGEList_plasma <- calcNormFactors(DGEList_plasma, method = "TMM")

```

## Filter biological noise - Plasma

Once the libraries are normalised we can filter for biological noise - i.e. miRNAs that are present in such small numbers that they are not biologically relevant. Here a count of less than 1 CPM in 7 or more samples is considered biologically relevant for this analysis.

```{r warning=FALSE, message=FALSE, error=FALSE}

# new df of raw cpm for the reduced DGEList
rawCpm_plasma <- cpm(DGEList_plasma)

# new df of unfiltered log 2 cpm for the reduced DGEList
rawlcpm_plasma <- cpm(DGEList_plasma, log = TRUE)

# remove low expressed genes logical eg cpm >1 in >=20 samples (7 samples in the smallest group)
x <- 1 # CPM for filtering
z <- 7 # number of samples for filtering

keep.exprs <- rowSums(rawCpm_plasma > x) >= z
DGEList_plasma <- DGEList_plasma[keep.exprs,, keep.lib.sizes = FALSE] 

```

# C19MC - Chromosome 14 miRNA mega cluster - *Placenta Specific Expression*

## Heatmap of C14MC placental miRNA in placenta - weeks' gestation as column names

```{r}

# set ggplot2 theme
theme_set(theme_bw())

cpmDGEList_placenta <- cpm(DGEList_placenta) %>%
  as.data.frame()

weeks <- samples[, c("samplename", "gestation")]

# all observed C14MC cluster miRNA in placenta
allC14MC_placenta <- subset(cpmDGEList_placenta, rownames(cpmDGEList_placenta) %in% c("hsa-miR-136-3p", "hsa-miR-136-5p", "hsa-miR-154-3p", "hsa-miR-154-5p", "hsa-miR-299-3p", "hsa-miR-299-5p", "hsa-miR-323a-3p", "hsa-miR-323a-5p", "hsa-miR-323b-3p", "hsa-miR-329-3p", "hsa-miR-329-5p", "hsa-miR-342-3p", "hsa-miR-342-5p", "hsa-miR-369-3p", "hsa-miR-369-5p", "hsa-miR-370-3p", "hsa-miR-370-5p", "hsa-miR-376a-3p", "hsa-miR-376a-5p", "hsa-miR-376b-3p", "hsa-miR-376b-5p", "hsa-miR-376c-3p", "hsa-miR-376c-5p", "hsa-miR-377-3p", "hsa-miR-377-5p", "hsa-miR-409-3p", "hsa-miR-409-5p", "hsa-miR-410-3p", "hsa-miR-410-5p", "hsa-miR-411-3p", "hsa-miR-411-5p", "hsa-miR-433-3p", "hsa-miR-433-5p", "hsa-miR-485-3p", "hsa-miR-485-5p", "hsa-miR-487a-3p", "hsa-miR-487a-5p", "hsa-miR-487b-3p", "hsa-miR-487b-5p", "hsa-miR-494-3p", "hsa-miR-494-5p","hsa-miR-495-3p", "hsa-miR-495-5p", "hsa-miR-654-3p", "hsa-miR-654-5p", "hsa-miR-1247-3p", "hsa-miR-1247-5p")) %>%
  as.data.frame() %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column() %>%
  rename(c('rowname' = 'samplename')) %>%
  left_join(., weeks, by = 'samplename') %>%
  select(., -samplename) %>%
  arrange(., gestation) %>%
  t()
nrow_placenta <- nrow(allC14MC_placenta)
nrow_placentaMinusOne <- nrow(allC14MC_placenta)-1
allC14MC_placenta <- allC14MC_placenta[c(nrow_placenta, 1:nrow_placentaMinusOne), ]
colnames(allC14MC_placenta) <-  allC14MC_placenta[1, ]
allC14MC_placenta <-  allC14MC_placenta[-1, ] 

# heat map - no annotation - includes weeks' as column names
pdf("C14MC.pdf", height = 11, width = 15)
pheatmap(as.matrix(allC14MC_placenta), 
         cluster_cols = T,
         cluster_rows = T,
         scale = "row",
         cellwidth = 15, 
         cellheight = 10,
         treeheight_row = 0,
         cutree_cols = 3,
         border_color = "white",
         main = "C14MC Expression in Placenta")
dev.off()

```

# Heat map C14MC miRNA with weeks as annotation - Placenta

```{r warning=FALSE, error=FALSE, message=FALSE}

lcpm_placenta <- cpm(DGEList_placenta, log = TRUE)

# all observed C14MC cluster miRNA in plasma
lcpm_placenta <- subset(lcpm_placenta, rownames(lcpm_placenta) %in% c("hsa-miR-136-3p", "hsa-miR-136-5p", "hsa-miR-154-3p", "hsa-miR-154-5p", "hsa-miR-299-3p", "hsa-miR-299-5p", "hsa-miR-323a-3p", "hsa-miR-323a-5p", "hsa-miR-323b-3p", "hsa-miR-329-3p", "hsa-miR-329-5p", "hsa-miR-342-3p", "hsa-miR-342-5p", "hsa-miR-369-3p", "hsa-miR-369-5p", "hsa-miR-370-3p", "hsa-miR-370-5p", "hsa-miR-376a-3p", "hsa-miR-376a-5p", "hsa-miR-376b-3p", "hsa-miR-376b-5p", "hsa-miR-376c-3p", "hsa-miR-376c-5p", "hsa-miR-377-3p", "hsa-miR-377-5p", "hsa-miR-409-3p", "hsa-miR-409-5p", "hsa-miR-410-3p", "hsa-miR-410-5p", "hsa-miR-411-3p", "hsa-miR-411-5p", "hsa-miR-433-3p", "hsa-miR-433-5p", "hsa-miR-485-3p", "hsa-miR-485-5p", "hsa-miR-487a-3p", "hsa-miR-487a-5p", "hsa-miR-487b-3p", "hsa-miR-487b-5p", "hsa-miR-494-3p", "hsa-miR-494-5p","hsa-miR-495-3p", "hsa-miR-495-5p", "hsa-miR-654-3p", "hsa-miR-654-5p", "hsa-miR-1247-3p", "hsa-miR-1247-5p")) %>%
  as.data.frame()


annotation <- data.frame(samples[, c(1, 3)]) %>%
  set_colnames(c("rowname", "gestation")) %>%
  as.data.frame() %>%
  tibble::column_to_rownames() 

#pdf("C14MCPlacenta.pdf", height = 11, width = 15)
pheatmap(as.matrix(lcpm_placenta), 
         annotation = annotation,
         cluster_cols = T,
         cluster_rows = T,
         Legend = T,
         cellwidth = 15, 
         cellheight = 10,
         show_colnames = F,
         show_rownames = T,
         treeheight_row = 0,
         cutree_cols = 3,
         border_color = "white",
         main = "C14MC Expression in Placenta")
#dev.off()


```

## heatmap of C14MC placental miRNA in plasma

```{r warning=FALSE, error=FALSE, message=FALSE}

# set ggplot2 theme
theme_set(theme_bw())

cpmDGEList_plasma <- cpm(DGEList_plasma) %>%
  as.data.frame()

weeks <- samples[, c("samplename", "gestation")]

# all observed C14MC miRNA in plasma
allC14MC_plasma <- subset(cpmDGEList_plasma, rownames(cpmDGEList_plasma) %in% c("hsa-miR-136-3p", "hsa-miR-136-5p", "hsa-miR-154-3p", "hsa-miR-154-5p", "hsa-miR-299-3p", "hsa-miR-299-5p", "hsa-miR-323a-3p", "hsa-miR-323a-5p", "hsa-miR-323b-3p", "hsa-miR-329-3p", "hsa-miR-329-5p", "hsa-miR-342-3p", "hsa-miR-342-5p", "hsa-miR-369-3p", "hsa-miR-369-5p", "hsa-miR-370-3p", "hsa-miR-370-5p", "hsa-miR-376a-3p", "hsa-miR-376a-5p", "hsa-miR-376b-3p", "hsa-miR-376b-5p", "hsa-miR-376c-3p", "hsa-miR-376c-5p", "hsa-miR-377-3p", "hsa-miR-377-5p", "hsa-miR-409-3p", "hsa-miR-409-5p", "hsa-miR-410-3p", "hsa-miR-410-5p", "hsa-miR-411-3p", "hsa-miR-411-5p", "hsa-miR-433-3p", "hsa-miR-433-5p", "hsa-miR-485-3p", "hsa-miR-485-5p", "hsa-miR-487a-3p", "hsa-miR-487a-5p", "hsa-miR-487b-3p", "hsa-miR-487b-5p", "hsa-miR-494-3p", "hsa-miR-494-5p","hsa-miR-495-3p", "hsa-miR-495-5p", "hsa-miR-654-3p", "hsa-miR-654-5p", "hsa-miR-1247-3p", "hsa-miR-1247-5p")) %>%
  as.data.frame() %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column() %>%
  rename(c('rowname' = 'samplename')) %>%
  left_join(., weeks, by = 'samplename') %>%
  select(., -samplename) %>%
  arrange(., gestation) %>%
  t()
nrow_plasma <- nrow(allC14MC_plasma)
nrow_plasmaMinusOne <- nrow(allC14MC_plasma)-1
allC14MC_plasma <- allC14MC_plasma[c(nrow_plasma, 1:nrow_plasmaMinusOne), ]
colnames(allC14MC_plasma) <-  allC14MC_plasma[1, ]
allC14MC_plasma <-  allC14MC_plasma[-1, ] 


#pdf("C14MCplasma.pdf", height = 11, width = 15)
pheatmap(as.matrix(allC14MC_plasma), 
         cluster_cols = T,
         cluster_rows = T,
         scale = "row",
         cellwidth = 15, 
         cellheight = 10,
         treeheight_row = 0,
         cutree_cols = 3,
         border_color = "white",
         main = "C14MC Expression in Maternal Plasma")
#dev.off()

```

# Heat map C14MC miRNA with weeks as annotation - Placenta

```{r warning=FALSE, error=FALSE, message=FALSE}

lcpm_plasma <- cpm(DGEList_plasma, log = TRUE)

# all observed C14MC cluster miRNA in plasma
lcpm_plasma <- subset(lcpm_plasma, rownames(lcpm_plasma) %in% c("hsa-miR-136-3p", "hsa-miR-136-5p", "hsa-miR-154-3p", "hsa-miR-154-5p", "hsa-miR-299-3p", "hsa-miR-299-5p", "hsa-miR-323a-3p", "hsa-miR-323a-5p", "hsa-miR-323b-3p", "hsa-miR-329-3p", "hsa-miR-329-5p", "hsa-miR-342-3p", "hsa-miR-342-5p", "hsa-miR-369-3p", "hsa-miR-369-5p", "hsa-miR-370-3p", "hsa-miR-370-5p", "hsa-miR-376a-3p", "hsa-miR-376a-5p", "hsa-miR-376b-3p", "hsa-miR-376b-5p", "hsa-miR-376c-3p", "hsa-miR-376c-5p", "hsa-miR-377-3p", "hsa-miR-377-5p", "hsa-miR-409-3p", "hsa-miR-409-5p", "hsa-miR-410-3p", "hsa-miR-410-5p", "hsa-miR-411-3p", "hsa-miR-411-5p", "hsa-miR-433-3p", "hsa-miR-433-5p", "hsa-miR-485-3p", "hsa-miR-485-5p", "hsa-miR-487a-3p", "hsa-miR-487a-5p", "hsa-miR-487b-3p", "hsa-miR-487b-5p", "hsa-miR-494-3p", "hsa-miR-494-5p","hsa-miR-495-3p", "hsa-miR-495-5p", "hsa-miR-654-3p", "hsa-miR-654-5p", "hsa-miR-1247-3p", "hsa-miR-1247-5p")) %>%
  as.data.frame()


annotation <- data.frame(samples[, c(1, 3)]) %>%
  set_colnames(c("rowname", "gestation")) %>%
  as.data.frame() %>%
  tibble::column_to_rownames() 

#pdf("C14MCPlasma.pdf", height = 11, width = 15)
pheatmap(as.matrix(lcpm_plasma), 
         annotation = annotation,
         cluster_cols = T,
         cluster_rows = T,
         Legend = T,
         cellwidth = 15, 
         cellheight = 10,
         show_colnames = F,
         show_rownames = T,
         treeheight_row = 0,
         cutree_cols = 3,
         border_color = "white",
         main = "C14MC Expression in Maternal Plasma")
#dev.off()


```

## geom_line of C14MC placental miRNA in placenta

```{r}

# mirbase <- mirbase_db[,7] %>%
#   as.data.frame() %>%
#   filter(., seqnames == "chr14") %>%
#   subset(., Name %in% c("hsa-miR-136-3p", "hsa-miR-136-5p", "hsa-miR-154-3p", "hsa-miR-154-5p", "hsa-miR-299-3p", "hsa-miR-299-5p", "hsa-miR-323a-3p", "hsa-miR-323a-5p", "hsa-miR-323b-3p", "hsa-miR-329-3p", "hsa-miR-329-5p", "hsa-miR-342-3p", "hsa-miR-342-5p", "hsa-miR-369-3p", "hsa-miR-369-5p", "hsa-miR-370-3p", "hsa-miR-370-5p", "hsa-miR-376a-3p", "hsa-miR-376a-5p", "hsa-miR-376b-3p", "hsa-miR-376b-5p", "hsa-miR-376c-3p", "hsa-miR-376c-5p", "hsa-miR-377-3p", "hsa-miR-377-5p", "hsa-miR-409-3p", "hsa-miR-409-5p", "hsa-miR-410-3p", "hsa-miR-410-5p", "hsa-miR-411-3p", "hsa-miR-411-5p", "hsa-miR-433-3p", "hsa-miR-433-5p", "hsa-miR-485-3p", "hsa-miR-485-5p", "hsa-miR-487a-3p", "hsa-miR-487a-5p", "hsa-miR-487b-3p", "hsa-miR-487b-5p", "hsa-miR-494-3p", "hsa-miR-494-5p","hsa-miR-495-3p", "hsa-miR-495-5p", "hsa-miR-654-3p", "hsa-miR-654-5p", "hsa-miR-1247-3p", "hsa-miR-1247-5p")) %>%
#   arrange(., Name)
#   
# mirbase <- mirbase[c(-4, -6, -10, -12, -19, -28), ] %>%
#   arrange(., start)
#   
# 
# mirbase$rank <- NA
# mirbase$rank <- 1:nrow(mirbase)
# 
# mirbase <- mirbase[, 6:7] %>%
#   rename(c('Name' = 'rowname'))
# 
# allC14MC_placenta <- as.data.frame(allC19MC_placenta) 
# allC14MC_placenta$rowname <- rownames(allC19MC_placenta) %>%
#   as.data.frame()
# 
# 
# mirbase <- left_join(mirbase, allC19MC_placenta, by = 'rowname')

```


```{r}

today <- Sys.Date()

```

This R session was run on `r format(today, format="%B %d %Y")`.

```{r}

sessionInfo()

```
