---
title: "placentaDifferentialExpression"
author: "Melanie Smith"
date: "20/09/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Import Counts and Sample Data

This step imports the raw counts data and samples metadata and reports some summary information.

```{r message=FALSE, warning=FALSE}

library(tidyverse)
library(pander)

# import miRNA counts table from VM
countsPlacenta <- read.table(("~/placentaBcbio/placenta/final/2017-05-23_placenta/counts_mirna.tsv"), 
                             sep="\t", 
                             header=T, 
                             row.names="mirna")

# remove the "_R1" and replace "Q" with "PAC" from the column names (eg Q006_S1_R1 becomes PAC006 to match the 'samples' data.frame) in the 'counts' data.frame
names(countsPlacenta) = sub("_R1","",names(countsPlacenta)) 
names(countsPlacenta) = sub("Q","PAC",names(countsPlacenta))
names(countsPlacenta) = sub("_S.*","",names(countsPlacenta))
# correct sample name mislabeled by QIAGEN
names(countsPlacenta) = sub("PAC033","PAC038",names(countsPlacenta))
# Reorder data.frame by columns
countsPlacenta <- countsPlacenta[,order(colnames(countsPlacenta),decreasing=FALSE)]

# import the samples data from VM 
samples <- readr::read_delim("~/placenta_miRNA.tsv", delim = "\t") %>% 
  dplyr::select(-starts_with("X"))

samples <- samples[c(1, 10, 12:17)] %>%
  as.data.frame()
names(samples) <- c("tempname", "biologicalGroup", "maternalAge", "gestation", "BMI", "smoker", "ethnicity", "boxPosition" )

# Remove the "+" from column 4 "Gestation" nb: escape special character using \\
samples$gestation <- gsub("\\+", "", samples$gestation)

# reassign object classes
samples$biologicalGroup <- as.factor(samples$biologicalGroup)
samples$gestation <- as.integer(samples$gestation)

# create a vector of just sample names from samples$samplename
tempname <- samples %>%
  dplyr::select(tempname)
# pad samplename in samples df to match count data df (PAC006 style)
samplename <- tempname %>% 
  mutate(samplename = paste0("PAC",stringr::str_pad(tempname, width = 3, pad = "0"))) %>%
  dplyr::select(samplename)
samplename <- as.data.frame(samplename)
samples <- c(samples, samplename)
# reorder columns (personal preference)
samples <- samples[c(9, 2, 3, 4, 5, 6, 7, 8, 1)]

# add new column named "group" into samples data.frame 
samples$group <- rep(NA, 48)

# add new group 'Trimester 1 or 2' factor to each row based on gestation
# Trimester 1 = 6-13 wks, Trimester 2 = 14-26 wks
samples$group <- ifelse(samples$gestation<=13, "Trimester_1", ifelse(samples$gestation>=14, "Trimester_2", NA))

# group as factor
samples$group <- as.factor(samples$group)

samples <- as.data.frame(samples)

# Add fetal sex to samples df
fetalSex <- read.csv("~/Bcbioplasma/fetalSex.csv", header = TRUE) %>%
  dplyr::select(-matches("gestation")) %>%
  mutate(samplename = paste0("PAC", stringr::str_pad(samplename, width = 3, pad = "0")))

fetalSex$Sex <- ifelse(fetalSex$fetalSex == "XX" , "Female", ifelse(fetalSex$fetalSex == "XY", "Male", NA))

samples <- left_join(samples, fetalSex, by = "samplename")

pandoc.table(samples[,c(-2, -8, -9, -11)],
             caption = "Meta Data for all placenta samples",
             split.table=Inf)

summary(samples$group)
summary(samples$fetalSex)


lapply(samples[, c(3:7)], function(x) {
    if (is.numeric(x)) return(summary(x))
    if (is.factor(x)) return(table(x))
})

```

### Create LaTeX code for the samples table in methods section

```{r warning=FALSE, error=FALSE}

library(xtable)
options(xtable.floating = FALSE)
options(xtable.timestamp = "")

# samples table for thesis
xtable(samples[,c(-2, -8,-9, -11)])

```

### Plot to visualise the pre filtered, un-normalised library sizes in Placenta

```{r warning=FALSE}

library(tidyverse)
library(magrittr)
library(dplyr)

# set ggplot2 theme
theme_set(theme_bw())

# count the number of 'non-zero' (ie expressed) miRNAs for each sample before filtering
nonzero <- function(x) sum(x != 0)
rawNonZero <- plyr::numcolwise(nonzero)(countsPlacenta)
rawNonZero <- t(rawNonZero) %>%
  as.data.frame()
rawNonZero <- tibble::rownames_to_column(rawNonZero) %>%
  set_colnames(c("samplename", "rawmiRNA")) %>%
  as.data.frame()

#png("rawmiRNAPlacenta.png")
ggplot(data=rawNonZero, aes(x=samplename, y=rawmiRNA, fill = samples$group)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Sample Name", y = "Count of Unique miRNAs Identified", title = "Count of pre-filtered miRNAs in Placenta") + 
  geom_hline(yintercept=1835,linetype="dashed") +
  geom_text(aes(13, 1900, label = "1835 Total miRNA"), size = 3, colour = "black") +
  scale_fill_discrete(name = "Trimester")
#dev.off()

```

We start with a total of 1835 unique miRNAs - of which no single sample has more than 1256 individual miRNAs. 

# Filter sequencing noise   

Here any individual observation with less than 5 counts is considered to be sequencing noise and is reduced to zero and removed by filtering.  
* Counts less than 5 reduced to 0
* Logic check and record how many miRs will be filtered
* Filter to remove miRs with zero counts in all samples

```{r warning=FALSE, message=FALSE}

# exclude any individual count less than 5
countsPlacenta[countsPlacenta < 5] <- 0

# check to establish if any miRNA have no expression across all samples
table(rowSums(countsPlacenta==0) == 48)

# remove miRNAs with zero counts in all samples
countsPlacenta <- countsPlacenta[ rowSums(countsPlacenta)!=0, ]

```

After filtering out the sequencing noise (sequence counts with low confidence) we are left with 1256 unique miRNAs.

# Establish DGEList object

The counts and samples data can now be combined into a single list object for use in differential expression analysis downstream.

```{r message=FALSE, warning=FALSE}

# load required Libraries
library(edgeR)
library(tidyverse)

# new df of miRNA names for DGEList object
genes <- as.character(rownames(countsPlacenta)) %>%
  as.data.frame() %>%
  set_colnames("SYMBOL") 

# new DGEList object for differential expression analysis
DGEList_placenta <- DGEList(counts=countsPlacenta, samples=samples, genes = genes)

# saveRDS(DGEList_placenta, "rawDGEList_placenta.rds")

```

# TMM normalisation of libraries

Before differential expression analysis the sample libraries need to be normalised to account for different library sizes.   
Normalising the libraries allows for the direct comparison between samples.   
Here the Trimmed Mean of M Values method is used.   
Prior to normalisation the placenta libraries are within a similar range, however the mean of counts is small and centered around 0-1 CPM (log2)   

```{r warning=FALSE}

library(RColorBrewer)

# Distribution of un-normalised counts 
nsamples <- ncol(DGEList_placenta)
col <- brewer.pal(nsamples, "Paired")

lcpm <- cpm(DGEList_placenta, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="Placenta: Un-normalised data",ylab="Log-cpm") 

# calculate normalisation factors and apply to the DGEList object
DGEList_placenta <- calcNormFactors(DGEList_placenta, method="TMM")

# saveRDS(DGEList_placenta, "normDGEList_placenta.rds")

# vector of normalisation factors if required
#DGEList_placenta$samples$norm.factors 

```

# Filter biological noise and plot 

Once the libraries are normalised we can filter for biological noise - i.e. miRNAs that are present in such small numbers that they are not biologically relevant. Here a count of less than 2 CPM in 7 or more samples is considered biologically relevant for this analysis. A count of 2 CPM is the equivalant of ~12 transcripts in the smallest library. The number of samples (here 7) is chosen such that it is possible for a miRNA to be expressed in every sample of the smallest group (here Trimester 2).   

```{r warning=FALSE, error=FALSE}

library(plyr)
library(tidyverse)
library(edgeR)

# set ggplot theme
theme_set(theme_bw())

# new df of unfiltered cpm for the reduced DGEList
rawCpm_placenta <- cpm(DGEList_placenta)

# new df of unfiltered log 2 cpm for the reduced DGEList
rawlcpm_placenta <- cpm(DGEList_placenta, log = TRUE)

# remove lowly expressed genes 
keep.exprs <- rowSums(rawCpm_placenta > 2) >= 7
DGEList_placenta <- DGEList_placenta[keep.exprs,, keep.lib.sizes=FALSE]

## The density of log-CPM values for pre-filtered data (A) and post-filtered data (B) are shown for each sample. 
## Dotted vertical lines mark the log-CPM of zero threshold (equivalent to a CPM value of 1) used in the filtering step.

nsamples <- ncol(DGEList_placenta)
col <- brewer.pal(nsamples, "Paired")
par(mfrow=c(1,2))
plot(density(rawlcpm_placenta[,1]), col=col[1], lwd=2, ylim=c(0,0.15), las=2,
  main="", xlab="")
title(main="A. Raw data", xlab="Log-cpm") 
abline(v=0, lty=3)
for (i in 2:nsamples){
  den <- density(rawlcpm_placenta[,i])
  lines(den$x, den$y, col=col[i], lwd=2)
}
#legend("topright", legend=samples$samplename, text.col=col, bty="n")

lcpm <- cpm(DGEList_placenta, log=TRUE)
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.15), las=2,
  main="", xlab="")
title(main="B. Filtered data", xlab="Log-cpm") 
abline(v=0, lty=3)
for (i in 2:nsamples){
  den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
#legend("topright", legend=samples$samplename, text.col=col, bty="n")

# plot of filtered and normalised library sizes
lcpm <- cpm(DGEList_placenta, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="B. Placenta: Normalised data",ylab="Log-cpm")

# Plot filtered library sizes against total miRNA
# count the number of 'non-zero' (ie expressed) miRNAs for each sample after filtering
filteredNonZero <- (DGEList_placenta[[1]]) %>%
  as.data.frame()

filteredNonZero <-  t(numcolwise(nonzero)(filteredNonZero)) %>%
  as.data.frame() %>%
  tibble::rownames_to_column()

#png("robustmiRNAPlacenta.png")
ggplot(data=filteredNonZero, aes(x=rowname, y=V1, fill=samples$group)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Sample Name", y = "Count of miRNAs", title = "Count Robust miRNAs in Placenta After Filtering") + 
  geom_hline(yintercept=714,linetype="dashed") +
  geom_text(aes(12, 750, label = "714 Total miRNA"), size = 3, colour = "black") +
  scale_fill_discrete(name = "Trimester")
#dev.off()

# Distribution of normalised and filtered counts data
lcpm <- cpm(DGEList_placenta, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="Placenta: Normalised data",ylab="Log-cpm")

```

```{r}
sessionInfo()
```